Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-14 MScDiplom_Lemenkova.tex
Comments 21 and 24 of Valentijn: updated 'Methods' chapter

The spatial resolution of \ac{Landsat ETM+} image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of \ac{ASTER} 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and \ac{NIR}, with spatial resolution of 1-4 meters. \ac{ASTER} and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork \textit{in-situ} measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.\\
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors \ac{Trios-RAMSES} Hyperspectral
\ac{UVA}/\ac{UVB} Irradiance Sensor and \ac{Trios-RAMSES}-ARC Hyperspectral \ac{UV}-\ac{VIS} Radiance Sensor (Fig. \ref{fig:2.1}), both adjusted for the measurements of the irradiance and radiance (see \nameref{app}: Tab. \ref{tab:A.1}, Tab. \ref{tab:A.2}). 
These products have been used for the radiance and irradiance measurements in at the \ac{HCMR} at the Institute of Aquaculture, Crete Island, 2009 by Ms Sylvia Noralez using following workflow. The spectrometer was adjusted for automatic measurements mode, with measurements taken as fast as possible. The spectrometer head was held submerged, and the sampling was controlled by an operator on the surface boat. The head of the sensor was pointed downward at an angle of $0\,^{\circ}$ (nadir) in order to capture the spatial discernability in the radiance for the benthic cover types. The frame was held at $45\,^{\circ}$ angle in order to keep sensor looking down at $0\,^{\circ}$ (nadir view). A waterproof camera was attached to the platform to assist with the identification of the target object  being measured (\cite{Noralez10}\label{Noralez10}).
The highest measured values are located in the diapason of 410-730 nm for the water irradiance (Fig. \ref{fig:A.18}) and 430-650 for the water radiance (Fig.\ref{fig:A.16}). 
Afterwards, the measured values of the radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:A.15}). The spectral range of radiance
cover diapason of 320-950 nm, and irradiance measurements are covered in the interval of 280-500
nm, which is suitable for characteristics of seagrass reflectance.
Different curves on the reflectance, radiance and irradiance graphs for example, on Fig.\ref{fig:A.12}, or \ref{fig:A.14}, enlarged) represent several series of the measurements. The values of the spectral reflectance are received from the
computations of these values using mathematical formulae. The graphs shown on Fig. \ref{fig:4.2}, Fig. \ref{fig:A.23}, Fig. \ref{fig:A.24}, Fig. \ref{fig:A.17} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and without sediments and suspended particles, respectively. Graph on Fig \ref{fig:A.13} displays statistical analysis of the measured sets of observations, i.e. shows the mid-spread of the statistical quartiles (Q1-Q3), mean and extreme values.

@@

\subsubsection{Implementation of statistical analysis}
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). \\
As a result, we received mean values of spectral reflectance of \textit{P.oceanica} for extreme, average, median and both quartiles,
independent of the individual sets 1-350. \\
We used several statistical methods to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.

@@

\subsubsection{Spectral simulation of aquatic objects}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.
The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.

\paragraph{Model parameters: depth and bottom cover fraction}
Although seagrass \textit{P.oceanica} can be found until depth limits down to 40 m depth \cite{DenHartog70}\label{DenHartog70}, the most preferable limits of its distribution in the Mediterranean Sea are shallow waters until
4 meters of depth.

@@

An analysis of the spectral reflectance of \textit{P.oceanica} is done using the \ac{WASI} simulations\index[concepts]{Seawater optical properties!RTM!WASI} in order to determine, which wavebands can be still used to identify \textit{P.oceanica}. The analysis of spectra shows
that the appropriate wavebands for seagrass mapping lay between 500 and 600 nm and has
also peaks at around 700 nm, ca between 680 and 710 nm (Fig.\ref{fig:4.10}). The highest values of the bottom
reflectance are at spectra of 500-600 nm. The most appropriate depths at which the spectral signatures of the seagrass
could still be discriminated are lesser than 2.5 meters.

\paragraph{Spectral discrimination of \textit{P.oceanica}}\label{page-39}
The patches of white sandy bottoms of the seafloor, studied in the fieldwork in Ligaria beach, are
much brighter than mattes of \textit{P.oceanica} (Fig.\ref{fig:A.6}, \ref{fig:A.5}): seafloor types in Ligaria), which can be clearly seen
at the graph comparing values of the spectral reflectance of the carbonate sand (Fig.\ref{fig:A.35}) and that of \textit{P.oceanica} (Fig.\ref{fig:A.28}).

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5, angle=90]{UML_StatAn.jpg}
	\caption{Work Flowchart Diagram illustrating algorithm of the Statistical Analysis applied towards the observation dataset (\href{http://staruml.sourceforge.net/en/}{StarUML}). Page: $<$\pageref{page-40}$>$}
	\label{fig:4.12}
\end{figure}

The method of \textit{least squares}\index[concepts]{Data!statistical analysis!least squares, method of} minimizes the sum of the squares of the errors made in solving every single observation made. We tried different approaches of the \textit{least squares} methods: Tab.\ref{tab:15}, Tab.\ref{tab:18}, Tab.\ref{tab:27}.\\
To focus on the relationship between a dependent variable (each single observation of the spectral reflectance) and independent variables (three various depths: 0.5, 1.5 and 2.5 meters) we used the \textit{regression analysis}, which enables modeling and analyzing values of the spectral reflectance at several depths: Tab.\ref{tab:5}, Tab.\ref{tab:6}, Tab.\ref{tab:7}. 
The summary of the values of spectral reflectance properties on various depths is presented on the Tab.\ref{tab:A.7}.
The \textit{k-means cluster analysis}\index[concepts]{Data!statistical analysis!k-means cluster analysis} has been used for partitioning observation sets into k clusters (Tab.\ref{tab:20}, Tab.\ref{tab:21}, Tab.\ref{tab:22}, Tab.\ref{tab:23}) in which each observation belongs to the cluster with the nearest mean.
It enables to highlight the main areas of the location of values of spectral reflectance along the spectra.\\
To estimate spectral density of a spectral reflectance measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30}), which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31}) of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12} ) 
and the frequencies we want to retain (the general profile that lies in the 450-850 nm, Fig.\ref{fig:A.14}, enlarged part of 500-660nm).\\
The visualization of the data plotting was made by means of Gnuplot software, which enables fine
drawing and advanced displaying of large amounts of serial data, ultimate control over graph properties, 
the simplicity of plotting and the ease of scripting (see Fig.\ref{fig:4.13}, \nameref{app}).
The graphs illustrating spectral signatures of various seafloor cover types display the mean values and
areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1, Q3, shaded areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in
spectra, where the measurements were done; the y-axis was adjusted for the better visualization of the
\textit{P.oceanica} spectra: as its spectral values are mostly located in the lower part of the spectra (usually no
more than 0.20 nm, except for borders), we did not extend the y-scale up to 100 percent, in order to
focus on the most necessary area of spectral values (Fig.\ref{fig:4.14}, Fig.\ref{fig:4.4} and \nameref{app}). The mean values are
highlighted using bold red points, as we used these values for plotting the final graph of \textit{P.oceanica} spectra in
Ligaria beach, Agia Pelagia.

>>>>

The spatial resolution of \ac{Landsat ETM+} image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of \ac{ASTER} 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and \ac{NIR}, with spatial resolution of 1-4 meters. \ac{ASTER} and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork \textit{in-situ} measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.\\
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors \ac{Trios-RAMSES} Hyperspectral
\ac{UVA}/\ac{UVB} Irradiance Sensor and \ac{Trios-RAMSES}-ARC Hyperspectral \ac{UV}-\ac{VIS} Radiance Sensor (Fig. \ref{fig:2.1}), both adjusted for the measurements of the irradiance and radiance (see \nameref{app}: Tab. \ref{tab:A.1}, Tab. \ref{tab:A.2}). 
These products have been used for the radiance and irradiance measurements in at the \ac{HCMR} at the Institute of Aquaculture, Crete Island, 2009 by Ms Sylvia Noralez using following workflow. The spectrometer was adjusted for automatic measurements mode, with measurements taken as fast as possible. The spectrometer head was held submerged, and the sampling was controlled by an operator on the surface boat. The head of the sensor was pointed downward at an angle of $0\,^{\circ}$ (nadir) in order to capture the spatial discernibility in the radiance for the benthic cover types. The frame was held at $45\,^{\circ}$ angle in order to keep sensor looking down at $0\,^{\circ}$ (nadir view). A waterproof camera was attached to the platform to assist with the identification of the target object  being measured (\cite{Noralez10}\label{Noralez10}).
The highest measured values are located in the diapason of 410-730 nm for the water irradiance (Fig. \ref{fig:A.18}) and 430-650 for the water radiance (Fig.\ref{fig:A.16}). 
Afterwards, the measured values of the radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:A.15}). The spectral range of radiance
cover diapason of 320-950 nm, and irradiance measurements are covered in the interval of 280-500
nm, which is suitable for characteristics of seagrass reflectance.
Different curves on the reflectance, radiance and irradiance graphs for example, on Fig.\ref{fig:A.12}, or \ref{fig:A.14}, enlarged) represent several series of the measurements. The values of the spectral reflectance are received from the
computations of these values using mathematical formulae. The graphs shown on Fig. \ref{fig:4.2}, Fig. \ref{fig:A.23}, Fig. \ref{fig:A.24}, Fig. \ref{fig:A.17} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and without sediments and suspended particles, respectively. Graph on Fig \ref{fig:A.13} displays statistical analysis of the measured sets of observations, i.e. shows the mid-spread of the statistical quartiles (Q1-Q3), mean and extreme values.

@@

\subsubsection{Implementation of statistical analysis}
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). \\
As a result, we received mean values of spectral reflectance of \textit{P.oceanica} for extreme, average, median and both quartiles,
independent of the individual sets 1-350. \\
We used several statistical methods (Fig.\ref{fig:4.12})\label{stan} to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.
\pagebreak

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5, angle=90]{UML_StatAn.jpg}
	\caption{Work Flowchart Diagram illustrating algorithm of the Statistical Analysis applied towards the observation dataset (\href{http://staruml.sourceforge.net/en/}{StarUML}). Page: $<$\pageref{page-40}$>$, $<$\pageref{stan}$>$}
	\label{fig:4.12}
\end{figure}

\subsection{Spectral simulation of aquatic objects}\index[concepts]{Data!processing}\label{page-35}
The main aim of this part of research is to clarify if the bottom reflectance of the different seafloor types including patches of
the seagrass \textit{P.oceanica} meadows, silt and carbonate sand differ and can be clearly discriminated while mapping. A study is
based on three seafloor types containing silt, carbonate sand and seagrass, as well as mixed
types, where the spectral signatures were examined. \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI} (Fig.\ref{fig:4.7}) is used to simulate
spectral reflectance and colour discrimination of water, affected by presence of \textit{P.oceanica} and other factors, under various
environmental conditions which influence its colour.
The data provided with the model was determined at freshwater of the Boddensee \cite{Gege05}\label{Gege05}, 
yet the model was adjusted for the marine environment, so that its parameters (concentration of chlorophyll, concentration of small particles,
yellow substance, etc) now perfectly simulate the Mediterranean water conditions.
The remote sensing reflectance has been compared under the conditions of different water depths and
cover fraction of the seafloor, in order to assess spectral signatures of the seagrass and carbonate sand
as major seafloor types. \ac{WASI} enables to use forward or inverse calculation for the spectrum types at
a diapason of 350-800 nm with a 1nm spectral resolution. For the spectral analysis we applied
forward calculations, i.e. a computing and plotting of series of spectra according to specified
parameter settings, with exactly defined depths and cover fraction.

\subsubsection{Model parameters: depth and bottom cover fraction}
Although seagrass \textit{P.oceanica} can be found until depth limits down to 40 m depth \cite{DenHartog70}\label{DenHartog70}, the most preferable limits of its distribution in the Mediterranean Sea are shallow waters until
4 meters of depth.

@@

\subsection{Spectral discrimination of \textit{P.oceanica}}

@@

An analysis of the spectral reflectance of \textit{P.oceanica} is done using the \ac{WASI} simulations\index[concepts]{Seawater optical properties!RTM!WASI} in order to determine, which wavebands can be still used to identify \textit{P.oceanica}. The analysis of spectra shows
that the appropriate wavebands for seagrass mapping lay between 500 and 600 nm and has
also peaks at around 700 nm, ca between 680 and 710 nm (Fig.\ref{fig:4.10}). The highest values of the bottom
reflectance are at spectra of 500-600 nm. The most appropriate depths at which the spectral signatures of the seagrass
could still be discriminated are lesser than 2.5 meters.
The patches of white sandy bottoms of the seafloor, studied in the fieldwork in Ligaria beach, are
much brighter than mattes of \textit{P.oceanica} (Fig.\ref{fig:A.6}, \ref{fig:A.5})\label{page-39}: seafloor types in Ligaria), which can be clearly seen
at the graph comparing values of the spectral reflectance of the carbonate sand (Fig.\ref{fig:A.35}) and that of \textit{P.oceanica} (Fig.\ref{fig:A.28}).

@@

The method of \textit{least squares}\index[concepts]{Data!statistical analysis!least squares, method of} minimizes the sum of the squares of the errors made in solving every single observation made. We tried different approaches of the \textit{least squares} methods: Tab.\ref{tab:15}, Tab.\ref{tab:18}, Tab.\ref{tab:27}.\\
To focus on the relationship between a dependent variable (each single observation of the spectral reflectance) and independent variables (three various depths: 0.5, 1.5 and 2.5 meters) we used the \textit{regression analysis}, which enables modeling and analyzing values of the spectral reflectance at several depths: Tab.\ref{tab:5}, Tab.\ref{tab:6}, Tab.\ref{tab:7}. 
The summary of the values of spectral reflectance properties on various depths is presented on the Tab.\ref{tab:A.7}.
The \textit{k-means cluster analysis}\index[concepts]{Data!statistical analysis!k-means cluster analysis} has been used for partitioning observation sets into k clusters (Tab.\ref{tab:20}, Tab.\ref{tab:21}, Tab.\ref{tab:22}, Tab.\ref{tab:23}) in which each observation belongs to the cluster with the nearest mean.
It enables to highlight the main areas of the location of values of spectral reflectance along the spectra.\\
To estimate spectral density of a spectral reflectance measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30}), which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31}) of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12} ) 
and the frequencies we want to retain (the general profile that lies in the 450-850 nm, Fig.\ref{fig:A.14}, enlarged part of 500-660nm).\\
The visualization of the data plotting was made by means of Gnuplot software, which enables fine
drawing and advanced displaying of large amounts of serial data, ultimate control over graph properties, 
the simplicity of plotting and the ease of scripting (see Fig.\ref{fig:4.13}, \nameref{app}).
The graphs illustrating spectral signatures of various seafloor cover types display the mean values and
areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1, Q3, shaded areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in
spectra, where the measurements were done; the y-axis was adjusted for the better visualization of the
\textit{P.oceanica} spectra: as its spectral values are mostly located in the lower part of the spectra (usually no
more than 0.20 nm, except for borders), we did not extend the y-scale up to 100 percent, in order to
focus on the most necessary area of spectral values (Fig.\ref{fig:4.14}, Fig.\ref{fig:4.4} and \nameref{app}). The mean values are
highlighted using bold red points, as we used these values for plotting the final graph of \textit{P.oceanica} spectra in
Ligaria beach, Agia Pelagia.
